---
title: "Homework2.rmd"
author: "Fabian & Samuel"
output:
  pdf_document: default
  html_document: default
date: "2024-12-09"
---

```{r setup, include=FALSE, tidy.opts=list(width.cutoff=60)}
### Packages and Data Setup

# Load the necessary packages
library(tidyverse)
library(nycflights13)
library(ggplot2)
library(dplyr)
library(ggridges)

```

## Task 1

```{r echo=FALSE}
ewr_cold_days <- weather %>%
filter(origin == "EWR") %>%
  mutate(temp_c = (temp - 32) * 5 / 9,      # Convert temperature to Celsius
         day = as.Date(time_hour)) %>%      # Extract the date
  filter(temp_c < 10) %>%                   # Filter for temperatures below 10°C
  group_by(month, day) %>%                  # Group by month and day
  summarise(n = n(), .groups = "drop") %>%  # Summarize daily grouping
  group_by(month) %>%                       # Group by month
  summarise(cold_days = n_distinct(day))    # Count unique cold days

# Display the results
print(ewr_cold_days)

# Check if dep_time contains NA values
flights %>%
  filter(is.na(dep_time)) %>%
  summarise(total_missing = n())

# Analyze flights in January with missing dep_time and other missing variables
missing_dep_time <- flights %>%
  filter(month == 1, is.na(dep_time)) %>%     # Filter January flights with missing dep_time
  summarise(
    total_missing_dep_time = n(),            # Number of affected flights
    across(everything(), ~ sum(is.na(.)))    # Count missing values for each column
  )

# Display the results
print(missing_dep_time)
```

## Task 2

```{r echo=FALSE}

# Pipeline to analyze flights with missing 'dep_time' in January
flights_missing_dep_time <- flights %>%
  filter(month == 1, is.na(dep_time)) %>%
  summarise(
    total_missing_dep_time = n(),                # Number of flights without departure time
    missing_sched_dep_time = sum(is.na(sched_dep_time)),  # Number of missing 'sched_dep_time' in these flights
    missing_arr_time = sum(is.na(arr_time)),      # Number of missing 'arr_time' in these flights
    missing_air_time = sum(is.na(air_time))       # Number of missing 'air_time' in these flights
  )

# Display the results
print(flights_missing_dep_time)

```

## Task 3 - a

```{r echo=FALSE}

# Example for creating 'weather_selected'
weather_selected <- weather %>%
  select(month, temp) %>%  # Select only the variables 'month' and 'temp'
  filter(!is.na(temp))     # Filter out all rows with missing 'temp' values

# a) Distribution of temperatures grouped by months with month names on the y-axis
ggplot(weather_selected, aes(x = temp)) +
  geom_histogram(binwidth = 1, color = "black", fill = "skyblue") +
  facet_wrap(~ factor(month, levels = 1:12, labels = month.name), scales = "free_y") +  # Use month.name for labels
  labs(x = expression(paste("Temperature (", degree, "F)")), y = "Count", title = "Temperature Distribution by Month") +  # Use expression()
  scale_x_continuous(breaks = seq(-20, 100, 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(strip.text = element_text(size = 12),
        axis.text = element_text(size = 10))

```

## Task 3 - b

```{r echo=FALSE}
# b) Density plot using ggridges::geom_density_ridges()
ggplot(weather_selected, aes(x = temp, y = factor(month))) +
  geom_density_ridges(scale = 3, fill = "lightblue", color = "blue") +  # Create ridge plots with specified colors and scaling
  labs(x = expression(paste("Temperature (", degree, "F)")), y = "Month", title = "Density Ridge Plot of Temperature by Month") +  # Add labels and title
  scale_y_discrete(labels = month.name) +  # Use month.name to label the y-axis
  theme_minimal()  # Apply a clean minimal theme

```

## Task 3 - c

```{r echo=FALSE}
# c) Aggregated monthly average temperature and standard deviation
aggregated_data <- weather_selected %>%
  group_by(month) %>%  # Group data by month
  summarise(
    mean_temp = mean(temp, na.rm = TRUE),  # Calculate the mean temperature for each month
    sd_temp = sd(temp, na.rm = TRUE),      # Calculate the standard deviation of the temperature
    n = n()                                # Count the number of observations per month
  ) %>%
  mutate(
    upper_bound = mean_temp + 3 * (sd_temp / sqrt(n)),  # Calculate the upper bound for the 99% confidence interval
    lower_bound = mean_temp - 3 * (sd_temp / sqrt(n))   # Calculate the lower bound for the 99% confidence interval
  )

# Plot for monthly average temperature with confidence intervals
ggplot(aggregated_data, aes(x = month, y = mean_temp)) +
  geom_line(color = "blue", size = 1) +  # Line plot for average temperature
  geom_ribbon(aes(ymin = lower_bound, ymax = upper_bound), fill = "lightblue", alpha = 0.5) +  # Confidence interval ribbon
  labs(
    x = "Month", 
    y = expression(paste("Average Temperature (", degree, "F)")), 
    title = "Monthly Average Temperature with 99% Confidence Interval"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.name) +  # Label months with their names
  theme_minimal() +  # Apply a clean minimal theme
  theme(axis.text = element_text(size = 10))  # Customize axis text size

```

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure2, echo=FALSE}
plot(pressure)
```

Note that the echo = FALSE parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### **Workload**

### We sat down at the beginning and divided up the tasks. The first thing we did was set up a Git repository so that we could easily collaborate and continuously track each other’s progress. We organized the tasks so that Samuel took on tasks 1, 2 and 3, while Fabian handled tasks 4 and 5. This clear division helped us focus and avoid overlapping efforts. After completing our initial assignments, we held a short meeting to discuss the status and share updates on our progress. This check-in proved beneficial, as it highlighted some areas that needed further refinement. Although a few items were still incomplete, we felt that we were moving in the right direction and understood what remained to be done. To ensure the quality of each other’s work, we decided that Fabian would review and make corrections to Samuel's tasks, and vice versa. This mutual review process not only helped catch errors but also facilitated a better understanding of each other's approach. By the end of the session, we felt more confident about our progress and looked forward to wrapping up the remaining tasks.
